---
title: ' '
date: "7/19/2022"
output:
  html_document:
    theme:
      bootswatch: minty
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries  
```{r, message=FALSE }
#tinytex::install_tinytex()
#library(tinytex)
library(magrittr)
library(dplyr)
#library(tidyverse)
library(picante)
library(geiger)
library(phytools)
library(ggplot2)
```

#### Filepaths and set wd
```{r }
setwd("C:/Users/Ali_Tew/OneDrive - University of North Carolina at Charlotte/Documents/R")
tree<-read.tree("notothenioid_timetree.tre")
eyeData<-read.csv("NotothenioidEyeSize.csv")
```
  
#### Read in data and summarize the traits
```{r }
#get the average measurement per species
updatedData<-eyeData %>% group_by(Name_in_tree) %>% mutate(
  meanEye=mean(c(Eye.Size,Eye.Size.1,Eye.Size.2), na.rm=TRUE),
  meanBodyLength=mean(c(Head.Length,Head.Length.1,Head.Length.2), na.rm=TRUE),
  meanBodyDepth=mean(c(Body.Depth,Body.Depth.1,Body.Depth.2), na.rm=TRUE))
```

#### Let's look at traits compared to the phylogeny
```{r, message=FALSE, warning=FALSE }
#match the tree and the trait of interest
eyedata<-updatedData$meanEye
names(eyedata)<-updatedData$Timetree_Name
td<-treedata(tree, eyedata)
eyeSize<-as.data.frame(td$data)

#subsample datasheet to remove anything not in tree
inTreeData<-updatedData[updatedData$Timetree_Name%in%rownames(eyeSize),]
```

#### pgls to see if mean eye diameter is correlated with mean body length
```{r, message=FALSE, warning=FALSE }
#first get object of taxa names for setting correlation structure
noto.names<-row.names(eyeSize)

#then set correlation structure for pgls
bm<-corBrownian(1, td$phy, form=~noto.names)

#now fit model
modelo1<-gls(meanEye~meanBodyLength, data=inTreeData, correlation=bm)
summary(modelo1)
inTreeData %>% ggplot(aes(x=log(meanBodyLength), y=meanEye)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#get residuals
resEye<-as.vector(modelo1$residuals)
names(resEye)<-inTreeData$Timetree_Name

#barplot with tree
plotTree.barplot(td$phy,resEye)
```

#### pgls to see if mean eye diameter is correlated with mean body depth
```{r, message=FALSE, warning=FALSE }
bm<-corBrownian(1, td$phy, form=~noto.names)
modelo2<-gls(meanEye~meanBodyDepth, data=inTreeData, correlation=bm)
summary(modelo2)
inTreeData %>% ggplot(aes(x=log(meanBodyDepth), y=meanEye)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#get residuals
resEye2<-as.vector(modelo2$residuals)
names(resEye2)<-inTreeData$Timetree_Name

#barplot with tree
plotTree.barplot(td$phy,resEye2)
```

#### pgls to see if residual eye size is correlated with residual body depth
```{r, message=FALSE, warning=FALSE }
#get residuals of eye size on standard length
bodyLength<-inTreeData$meanBodyLength
Eye<-inTreeData$meanEye
eye.fit<-glm(bodyLength~Eye)
eye.res<-residuals(eye.fit)

#get residuals of body depth on standard length
bodyDepth<-inTreeData$meanBodyDepth
depth.fit<-glm(bodyLength~bodyDepth)
depth.res<-residuals(depth.fit)

res.data<-as.data.frame(cbind(eye.res,depth.res))
row.names(res.data)<-row.names(eyeSize)

#fit model

bm<-corBrownian(1, td$phy, form=~noto.names)
modelo3<-gls(eye.res~depth.res, data=res.data, correlation=bm)
summary(modelo3)
res.data %>% ggplot(aes(x=log(depth.res), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")
```

#### pgls if residual eye size is correlated with depth (avg for main text, plots of mean, min, and max for supplemental materials)
```{r, message=FALSE, warning=FALSE }
#avg depth
avgDepth<-as.data.frame(cbind(eye.res,inTreeData$AMLR1998.2018_AvgDepth))
row.names(avgDepth)<-row.names(eyeSize)
avgDepth<-na.omit(avgDepth)
colnames(avgDepth)[2]<-"average.depth"

avgDepth.tax<-row.names(avgDepth)
avgDepth.tree<-keep.tip(tree,avgDepth.tax)

bm<-corBrownian(1, avgDepth.tree, form=~avgDepth.tax)
modelo4<-gls(eye.res~average.depth, data=avgDepth, correlation=bm)
summary(modelo4)
avgDepth %>% ggplot(aes(x=log(average.depth), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#min depth
minDepth<-as.data.frame(cbind(eye.res,inTreeData$Eastman2017_MinDepth))
row.names(minDepth)<-row.names(eyeSize)
minDepth<-na.omit(minDepth)
colnames(minDepth)[2]<-"min.depth"

minDepth.tax<-row.names(minDepth)
minDepth.tree<-keep.tip(tree,minDepth.tax)

bm<-corBrownian(1, minDepth.tree, form=~minDepth.tax)
modelo5<-gls(eye.res~min.depth, data=minDepth, correlation=bm)
summary(modelo5)
minDepth %>% ggplot(aes(x=log(min.depth), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#max depth
maxDepth<-as.data.frame(cbind(eye.res,inTreeData$Eastman2017_MaxDepth))
row.names(maxDepth)<-row.names(eyeSize)
maxDepth<-na.omit(maxDepth)
colnames(maxDepth)[2]<-"max.depth"

maxDepth.tax<-row.names(maxDepth)
maxDepth.tree<-keep.tip(tree,maxDepth.tax)

bm<-corBrownian(1, maxDepth.tree, form=~maxDepth.tax)
modelo6<-gls(eye.res~max.depth, data=maxDepth, correlation=bm)
summary(modelo6)
maxDepth %>% ggplot(aes(x=log(max.depth), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")
```

#### pgls if residual eye size is correlated with buoyancy
```{r, message=FALSE, warning=FALSE }
#avg buoyancy - proxy for water column niche use
MeanBuoy<-as.data.frame(cbind(eye.res,inTreeData$MeanBuoy))
row.names(MeanBuoy)<-row.names(eyeSize)
MeanBuoy<-na.omit(MeanBuoy)
colnames(MeanBuoy)[2]<-"avg.buoy"

MeanBuoy.tax<-row.names(MeanBuoy)
MeanBuoy.tree<-keep.tip(tree,MeanBuoy.tax)

bm<-corBrownian(1, MeanBuoy.tree, form=~MeanBuoy.tax)
modelo7<-gls(eye.res~avg.buoy, data=MeanBuoy, correlation=bm)
summary(modelo7)
MeanBuoy %>% ggplot(aes(x=log(avg.buoy), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

```

#### generate a phylomorphospace of the above
```{r, message=FALSE, warning=FALSE }
#prepare dataframe
all.traits<-cbind(inTreeData$AMLR1998.2018_AvgDepth,inTreeData$Eastman2017_MinDepth,inTreeData$Eastman2017_MaxDepth,inTreeData$MeanBuoy,res.data)
rownames(all.traits)<-rownames(eyeSize)
complete.traits<-na.omit(all.traits)

#prune tree to match data
PCA.tips<-row.names(complete.traits)
pPCA.tree<-keep.tip(tree,PCA.tips)

#conduct phylo pca
noto.pPCA<-phyl.pca(pPCA.tree,complete.traits,mode="cov")
attributes(noto.pPCA)
noto.pPCA$V
pPCA.PCscores<-noto.pPCA$S
row.names(pPCA.PCscores)<-row.names(complete.traits)

#or conduct regular pca
noto.PCA<-prcomp(complete.traits)
#get loadings
aload2 <- abs(noto.PCA$rotation)
sweep(aload2, 2, colSums(aload2), "/")
PCA.PCscores<-noto.PCA$x
row.names(PCA.PCscores)<-row.names(complete.traits)

#make contmap with color coded branches
AA<-contMap(pPCA.tree, pPCA.PCscores[,1])

# check to verify that temporal information is correct
phylomorphospace(AA$tree,pPCA.PCscores[,c(1,2)],colors=AA$cols,lwd=3, node.by.map=TRUE,xlab="PC 1",ylab="PC 2 ")
```

#### plot a phenogram of ecology reconstructed on the tree with residual eye size
```{r, message=FALSE, warning=FALSE }
#average depth
#data manipulation to get traits of interest
depths<-avgDepth$average.depth
names(depths)<-row.names(avgDepth)

eyes<-avgDepth$eye.res
names(eyes)<-row.names(avgDepth)

#make contmap with branches color-coded to reflect average depth
depthTree<-contMap(avgDepth.tree, depths)

#plot phenogram
phenogram(depthTree$tree, eyes, colors=depthTree$cols)

#minimum depth
#data manipulation to get traits of interest
min.depths<-minDepth$min.depth
names(min.depths)<-row.names(minDepth)

min.eyes<-minDepth$eye.res
names(min.eyes)<-row.names(minDepth)

#make contmap with branches color-coded to reflect average depth
min.depthTree<-contMap(minDepth.tree, min.depths)

#plot phenogram
phenogram(min.depthTree$tree, min.eyes, colors=min.depthTree$cols)

#maximum depth
#data manipulation to get traits of interest
max.depths<-maxDepth$max.depth
names(max.depths)<-row.names(maxDepth)

max.eyes<-maxDepth$eye.res
names(max.eyes)<-row.names(maxDepth)

#make contmap with branches color-coded to reflect average depth
max.depthTree<-contMap(maxDepth.tree, max.depths)

#plot phenogram
phenogram(max.depthTree$tree, max.eyes, colors=max.depthTree$cols)

#average buoyancy
#data manipulation to get traits of interest
avg.buoy<-MeanBuoy$avg.buoy
names(avg.buoy)<-row.names(MeanBuoy)

buoy.eyes<-MeanBuoy$eye.res
names(buoy.eyes)<-row.names(MeanBuoy)

#make contmap with branches color-coded to reflect average depth
avg.buoyTree<-contMap(MeanBuoy.tree, avg.buoy)

#plot phenogram
phenogram(avg.buoyTree$tree, buoy.eyes, colors=avg.buoyTree$cols)

```

#### disparity through time analyses for residual eye size, depth, and buoyancy
```{r, message=FALSE, warning=FALSE }
#create unique variables for each of the traits and omit species with missing data for traits
eye.size<-as.matrix(eye.res)
row.names(eye.size)<-inTreeData$Timetree_Name
#remove species with missing data
eye.size<-na.omit(eye.size)

depth<-as.matrix(inTreeData$AMLR1998.2018_AvgDepth)
row.names(depth)<-inTreeData$Timetree_Name
depth<-na.omit(depth)

buoy<-as.matrix(inTreeData$MeanBuoy)
row.names(buoy)<-inTreeData$Timetree_Name
buoy<-na.omit(buoy)

####perform disparity through time analyses for each trait 

#first prune tree such that tips match species for which we have trait data
eye.tax<-row.names(eye.size)
eye.tree<-keep.tip(tree, eye.tax)

#perform disparity through time analysis, calculating Morphological Disparity Index for only the first 80% of the phylogeny
dtt_eyeSize<-dtt(phy=eye.tree, data=eye.size, mdi.range=c(0.2,1), nsim=10000, plot=TRUE, calculateMDIp = TRUE)
dtt_eyeSize$MDI
dtt_eyeSize$MDIpVal


##depth

depth.tax<-row.names(depth)
depth.tree<-keep.tip(tree, depth.tax)

dtt_depth<-dtt(phy=depth.tree, data=depth, mdi.range=c(0.2,1), nsim=10000, plot=TRUE, calculateMDIp = TRUE)
dtt_depth$MDI
dtt_depth$MDIpVal


##size

buoy.tax<-row.names(buoy)
buoy.tree<-keep.tip(tree, buoy.tax)

dtt_buoy<-dtt(phy=buoy.tree, data=buoy, mdi.range=c(0.2,1), nsim=10000, plot=TRUE, calculateMDIp = TRUE)
dtt_buoy$MDI
dtt_buoy$MDIpVal

```

#### calculating phylogenetic signal in traits
```{r, message=FALSE, warning=FALSE }
###eye size

#using Blomberg et al's K

eyeK<-phylosig(eye.tree, eye.size, method="K", nsim=1000)
eyeK

eyeK.test<-phylosig(eye.tree, eye.size, method="K", test=TRUE, nsim=10000)
eyeK.test

plot(eyeK.test)

#using Pagel's lambda

eyeLambda<-phylosig(eye.tree, eye.size, method = "lambda")
eyeLambda

eyeLambda.test<-phylosig(eye.tree, eye.size, method = "lambda", test=TRUE)
eyeLambda.test

plot(eyeLambda.test)

###depth

#using Blomberg et al's K

depthK<-phylosig(depth.tree, depth, method="K", nsim=1000)
depthK

depthK.test<-phylosig(depth.tree, depth, method="K", test=TRUE, nsim=10000)
depthK.test

plot(depthK.test)

#using Pagel's lambda

depthLambda<-phylosig(depth.tree, depth, method = "lambda")
depthLambda

depthLambda.test<-phylosig(depth.tree, depth, method = "lambda", test=TRUE)
depthLambda.test

plot(depthLambda.test)


###buoyancy

#using Blomberg's K

#using Blomberg et al's K

buoyK<-phylosig(buoy.tree, buoy, method="K", nsim=1000)
buoyK

buoyK.test<-phylosig(buoy.tree, buoy, method="K", test=TRUE, nsim=10000)
buoyK.test

plot(buoyK.test)

#using Pagel's lambda

buoyLambda<-phylosig(buoy.tree, buoy, method = "lambda")
buoyLambda

buoyLambda.test<-phylosig(buoy.tree, buoy, method = "lambda", test=TRUE)
buoyLambda.test

plot(buoyLambda.test)

```







