---
title: ' '
date: "4/28/2024"
output:
  html_document:
    theme:
      bootswatch: minty
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries  
```{r, message=FALSE }
#tinytex::install_tinytex()
library(tinytex)
library(caper)
library(tidyverse)
library(picante)
library(geiger)
library(phytools)
library(l1ou)
library(motmot)

```

#### Filepaths and set wd
```{r }
setwd("~/Documents/Nototheniod_opsin/Revision_Analysis")
phylogeny<-"~/Documents/Nototheniod_opsin/notothenioid_timetree.tre"
data<-"~/Documents/Nototheniod_opsin/NotothenioidEyeSize.csv"
```
  
#### Read in data and summarize the traits
```{r }
#read in data
tree<-read.tree(phylogeny)
eyeData<-read.csv(data)
```

```{r }
#get the average measurement per species
updatedData<-eyeData %>% group_by(Name_in_tree) %>% mutate(
  meanEye=mean(c(Eye.Size,Eye.Size.1,Eye.Size.2), na.rm=TRUE),
  sdEye=sd(c(Eye.Size,Eye.Size.1,Eye.Size.2), na.rm=TRUE),
  meanBodyLength=mean(c(Standard.Length,Standard.Length.1,Standard.Length.2), na.rm=TRUE),
  sdBodyLength=sd(c(Standard.Length,Standard.Length.1,Standard.Length.2), na.rm=TRUE),
  meanBodyDepth=mean(c(Body.Depth,Body.Depth.1,Body.Depth.2), na.rm=TRUE),
  sdBodyDepth=sd(c(Body.Depth,Body.Depth.1,Body.Depth.2), na.rm=TRUE),
  meanHeadSize=(mean(c(Head.Length,Head.Length.1,Head.Length.2),na.rm=TRUE)),
  sdHeadSize=(sd(c(Head.Length,Head.Length.1,Head.Length.2),na.rm=TRUE))
  )

#plot the average sizes
updatedData<-unite(updatedData, merged_name, Genus, Species, sep = "_", remove = FALSE)

plot_body_length <- ggplot(updatedData, aes(merged_name, meanBodyLength)) +
  geom_bar(stat = "identity", fill = "blue", color = "black") +
  geom_errorbar(aes(ymin = meanBodyLength - sdBodyLength, ymax = meanBodyLength + sdBodyLength), width = 0.2) +
  coord_flip() +
  labs(x = "Fish Species", y = "Average Body Length (cm)") +
  theme_minimal()

plot_eye_size <- ggplot(updatedData, aes(merged_name, meanEye)) +
  geom_bar(stat = "identity", fill = "blue", color = "black") +
  geom_errorbar(aes(ymin = meanEye - sdEye, ymax = meanEye + sdEye), width = 0.2) +
  coord_flip() +
  labs(x = "Fish Species", y = "Average Eye Size (cm)") +
  theme_minimal()
```

#### Let's look at traits compared to the phylogeny
```{r, message=FALSE, warning=FALSE }
#match the tree and the trait of interest
eyedata<-updatedData$meanEye
names(eyedata)<-updatedData$Timetree_Name
td<-treedata(tree, eyedata)
eyeSize<-as.data.frame(td$data)

#subsample datasheet to remove anything not in tree
inTreeData<-updatedData[updatedData$Timetree_Name%in%rownames(eyeSize),]
```

#### pgls to see if mean eye diameter is correlated with mean body length
```{r, message=FALSE, warning=FALSE }
#first get object of taxa names for setting correlation structure
row.names(eyeSize)->noto.names

#then set correlation structure for pgls
bm<-corBrownian(1, td$phy, form=~noto.names)

#now fit model
modelo1<-gls(meanEye~meanBodyLength, data=inTreeData, correlation=bm)
summary(modelo1)
inTreeData %>% ggplot(aes(x=log(meanBodyLength), y=meanEye)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#get residuals
resEye<-as.vector(modelo1$residuals)
names(resEye)<-inTreeData$Timetree_Name

#barplot with tree
plotTree.barplot(td$phy,resEye)
```

#### pgls to see if mean eye diameter is correlated with mean body depth
```{r, message=FALSE, warning=FALSE }
bm<-corBrownian(1, td$phy, form=~noto.names)
modelo2<-gls(meanEye~meanBodyDepth, data=inTreeData, correlation=bm)
summary(modelo2)
inTreeData %>% ggplot(aes(x=log(meanBodyDepth), y=meanEye)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#get residuals
resEye2<-as.vector(modelo2$residuals)
names(resEye2)<-inTreeData$Timetree_Name

#barplot with tree
plotTree.barplot(td$phy,resEye2)
```

#### pgls to see if residual eye size is correlated with residual body depth
```{r, message=FALSE, warning=FALSE }
#get residuals of eye size on standard length
inTreeData$meanBodyLength->bodyLength
inTreeData$meanEye->Eye
glm(bodyLength~Eye)->eye.fit
residuals(eye.fit)->eye.res

#get residuals of body depth on standard length
inTreeData$meanBodyDepth->bodyDepth
glm(bodyLength~bodyDepth)->depth.fit
residuals(depth.fit)->depth.res

as.data.frame(cbind(eye.res,depth.res))->res.data
row.names(eyeSize)->row.names(res.data)

#fit model

bm<-corBrownian(1, td$phy, form=~noto.names)
modelo3<-gls(eye.res~depth.res, data=res.data, correlation=bm)
summary(modelo3)
res.data %>% ggplot(aes(x=log(depth.res), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")
```

#### pgls if residual eye size is correlated with depth (avg for main text, plots of mean, min, and max for supplemental materials)
```{r, message=FALSE, warning=FALSE }
#avg depth
as.data.frame(cbind(eye.res,inTreeData$AMLR1998.2018_AvgDepth))->avgDepth
row.names(eyeSize)->row.names(avgDepth)
na.omit(avgDepth)->avgDepth
colnames(avgDepth)[2]<-"average.depth"

row.names(avgDepth)->avgDepth.tax
avgDepth.tree<-keep.tip(tree,avgDepth.tax)

bm<-corBrownian(1, avgDepth.tree, form=~avgDepth.tax)
modelo4<-gls(eye.res~average.depth, data=avgDepth, correlation=bm)
summary(modelo4)
avgDepth %>% ggplot(aes(x=log(average.depth), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#min depth
as.data.frame(cbind(eye.res,inTreeData$Eastman2017_MinDepth))->minDepth
row.names(eyeSize)->row.names(minDepth)
na.omit(minDepth)->minDepth
colnames(minDepth)[2]<-"min.depth"

row.names(minDepth)->minDepth.tax
minDepth.tree<-keep.tip(tree,minDepth.tax)

bm<-corBrownian(1, minDepth.tree, form=~minDepth.tax)
modelo5<-gls(eye.res~min.depth, data=minDepth, correlation=bm)
summary(modelo5)
minDepth %>% ggplot(aes(x=log(min.depth), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

#max depth
as.data.frame(cbind(eye.res,inTreeData$Eastman2017_MaxDepth))->maxDepth
row.names(eyeSize)->row.names(maxDepth)
na.omit(maxDepth)->maxDepth
colnames(maxDepth)[2]<-"max.depth"

row.names(maxDepth)->maxDepth.tax
maxDepth.tree<-keep.tip(tree,maxDepth.tax)

bm<-corBrownian(1, maxDepth.tree, form=~maxDepth.tax)
modelo6<-gls(eye.res~max.depth, data=maxDepth, correlation=bm)
summary(modelo6)
maxDepth %>% ggplot(aes(x=log(max.depth), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")
```

#### pgls if residual eye size is correlated with buoyancy
```{r, message=FALSE, warning=FALSE }
#avg buoyancy - proxy for water column niche use
as.data.frame(cbind(eye.res,inTreeData$MeanBuoy))->MeanBuoy
row.names(eyeSize)->row.names(MeanBuoy)
na.omit(MeanBuoy)->MeanBuoy
colnames(MeanBuoy)[2]<-"avg.buoy"

row.names(MeanBuoy)->MeanBuoy.tax
MeanBuoy.tree<-keep.tip(tree,MeanBuoy.tax)

bm<-corBrownian(1, MeanBuoy.tree, form=~MeanBuoy.tax)
modelo7<-gls(eye.res~avg.buoy, data=MeanBuoy, correlation=bm)
summary(modelo7)
MeanBuoy %>% ggplot(aes(x=log(avg.buoy), y=eye.res)) + 
  geom_point() +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred", fill="blue")

```

#### generate a phylomorphospace of the above
```{r, message=FALSE, warning=FALSE }
#prepare dataframe
cbind(inTreeData$AMLR1998.2018_AvgDepth,inTreeData$Eastman2017_MinDepth,inTreeData$Eastman2017_MaxDepth,inTreeData$MeanBuoy,res.data)->all.traits
rownames(all.traits)<-rownames(eyeSize)
na.omit(all.traits)->complete.traits

#prune tree to match data
row.names(complete.traits)->PCA.tips
pPCA.tree<-keep.tip(tree,PCA.tips)

#conduct phylo pca
phyl.pca(pPCA.tree,complete.traits,mode="cov")->noto.pPCA
attributes(noto.pPCA)
noto.pPCA$V
noto.pPCA$S->pPCA.PCscores
row.names(pPCA.PCscores)<-row.names(complete.traits)

#or conduct regular pca
prcomp(complete.traits)->noto.PCA
#get loadings
aload2 <- abs(noto.PCA$rotation)
sweep(aload2, 2, colSums(aload2), "/")
noto.PCA$x->PCA.PCscores
row.names(PCA.PCscores)<-row.names(complete.traits)

#make contmap with color coded branches
AA<-contMap(pPCA.tree, pPCA.PCscores[,1])

# check to verify that temporal information is correct
phylomorphospace(AA$tree,pPCA.PCscores[,c(1,2)],colors=AA$cols,lwd=3, node.by.map=TRUE,xlab="PC 1",ylab="PC 2 ")
```

#### plot a phenogram of ecology reconstructed on the tree with residual eye size
```{r, message=FALSE, warning=FALSE }
#average depth
#data manipulation to get traits of interest
avgDepth$average.depth->depths
names(depths)<-row.names(avgDepth)

avgDepth$eye.res->eyes
names(eyes)<-row.names(avgDepth)

#make contmap with branches color-coded to reflect average depth
depthTree<-contMap(avgDepth.tree, depths)

#plot phenogram
phenogram(depthTree$tree, eyes, colors=depthTree$cols)

#minimum depth
#data manipulation to get traits of interest
minDepth$min.depth->min.depths
names(min.depths)<-row.names(minDepth)

minDepth$eye.res->min.eyes
names(min.eyes)<-row.names(minDepth)

#make contmap with branches color-coded to reflect average depth
min.depthTree<-contMap(minDepth.tree, min.depths)

#plot phenogram
phenogram(min.depthTree$tree, min.eyes, colors=min.depthTree$cols)

#maximum depth
#data manipulation to get traits of interest
maxDepth$max.depth->max.depths
names(max.depths)<-row.names(maxDepth)

maxDepth$eye.res->max.eyes
names(max.eyes)<-row.names(maxDepth)

#make contmap with branches color-coded to reflect average depth
max.depthTree<-contMap(maxDepth.tree, max.depths)

#plot phenogram
phenogram(max.depthTree$tree, max.eyes, colors=max.depthTree$cols)

#average buoyancy
#data manipulation to get traits of interest
MeanBuoy$avg.buoy->avg.buoy
names(avg.buoy)<-row.names(MeanBuoy)

MeanBuoy$eye.res->buoy.eyes
names(buoy.eyes)<-row.names(MeanBuoy)

#make contmap with branches color-coded to reflect average depth
avg.buoyTree<-contMap(MeanBuoy.tree, avg.buoy)

#plot phenogram
phenogram(avg.buoyTree$tree, buoy.eyes, colors=avg.buoyTree$cols)

```

#### disparity through time analyses for residual eye size, depth, and buoyancy
```{r, message=FALSE, warning=FALSE }
#create unique variables for each of the traits and omit species with missing data for traits
eye.size<-as.matrix(eye.res)
row.names(eye.size)<-inTreeData$Timetree_Name
#remove species with missing data
na.omit(eye.size)->eye.size

depth<-as.matrix(inTreeData$AMLR1998.2018_AvgDepth)
row.names(depth)<-inTreeData$Timetree_Name
na.omit(depth)->depth

buoy<-as.matrix(inTreeData$MeanBuoy)
row.names(buoy)<-inTreeData$Timetree_Name
na.omit(buoy)->buoy

####perform disparity through time analyses for each trait 

#first prune tree such that tips match species for which we have trait data
row.names(eye.size)->eye.tax
eye.tree<-keep.tip(tree, eye.tax)

#perform disparity through time analysis, calculating Morphological Disparity Index for only the first 80% of the phylogeny
dtt_eyeSize<-dtt(phy=eye.tree, data=eye.size, mdi.range=c(0.2,1), nsim=10000, plot=TRUE, calculateMDIp = TRUE)
dtt_eyeSize$MDI
dtt_eyeSize$MDIpVal


##depth

row.names(depth)->depth.tax
depth.tree<-keep.tip(tree, depth.tax)

dtt_depth<-dtt(phy=depth.tree, data=depth, mdi.range=c(0.2,1), nsim=10000, plot=TRUE, calculateMDIp = TRUE)
dtt_depth$MDI
dtt_depth$MDIpVal


##size

row.names(buoy)->buoy.tax
buoy.tree<-keep.tip(tree, buoy.tax)

dtt_buoy<-dtt(phy=buoy.tree, data=buoy, mdi.range=c(0.2,1), nsim=10000, plot=TRUE, calculateMDIp = TRUE)
dtt_buoy$MDI
dtt_buoy$MDIpVal

```

#### calculating phylogenetic signal in traits
```{r, message=FALSE, warning=FALSE }
###eye size

#using Blomberg et al's K

eyeK<-phylosig(eye.tree, eye.size, method="K", nsim=1000)
eyeK

eyeK.test<-phylosig(eye.tree, eye.size, method="K", test=TRUE, nsim=10000)
eyeK.test

plot(eyeK.test)

#using Pagel's lambda

eyeLambda<-phylosig(eye.tree, eye.size, method = "lambda")
eyeLambda

eyeLambda.test<-phylosig(eye.tree, eye.size, method = "lambda", test=TRUE)
eyeLambda.test

plot(eyeLambda.test)

###depth

#using Blomberg et al's K

depthK<-phylosig(depth.tree, depth, method="K", nsim=1000)
depthK

depthK.test<-phylosig(depth.tree, depth, method="K", test=TRUE, nsim=10000)
depthK.test

plot(depthK.test)

#using Pagel's lambda

depthLambda<-phylosig(depth.tree, depth, method = "lambda")
depthLambda

depthLambda.test<-phylosig(depth.tree, depth, method = "lambda", test=TRUE)
depthLambda.test

plot(depthLambda.test)


###buoyancy

#using Blomberg's K

#using Blomberg et al's K

buoyK<-phylosig(buoy.tree, buoy, method="K", nsim=1000)
buoyK

buoyK.test<-phylosig(buoy.tree, buoy, method="K", test=TRUE, nsim=10000)
buoyK.test

plot(buoyK.test)

#using Pagel's lambda

buoyLambda<-phylosig(buoy.tree, buoy, method = "lambda")
buoyLambda

buoyLambda.test<-phylosig(buoy.tree, buoy, method = "lambda", test=TRUE)
buoyLambda.test

plot(buoyLambda.test)

```

#### calculating phylogenetic signal in traits
```{r, message=FALSE, warning=FALSE }
###ANCOVA

#then set correlation structure for pgls
bm<-corBrownian(1, td$phy, form=~noto.names)

#see if the niche effects scaling
ANCOVA1<-gls(meanEye~meanBodyLength+Eastman2020_niche, data=inTreeData, correlation=bm)
summary(ANCOVA1)

#see if the niche effects scaling with body depth
ANCOVA2<-gls(meanEye~meanBodyDepth+Eastman2020_niche, data=inTreeData, correlation=bm)
summary(ANCOVA2)

#see if the niche effects scaling with headlength
ANCOVA3<-gls(meanEye~meanHeadSize+Eastman2020_niche, data=inTreeData, correlation=bm)
summary(ANCOVA3)
```

### OU Model comparison

#### pgls to see if mean eye diameter is correlated with mean body length
```{r, message=FALSE, warning=FALSE }
#then set correlation structure for pgls
ou<-corPagel(1, td$phy, form=~noto.names, fixed = FALSE)

#now fit model
OUmodelo1<-gls(meanEye~meanBodyLength, data=inTreeData, correlation=ou)
summary(OUmodelo1)
AIC(modelo1,OUmodelo1)

```

#### pgls to see if mean eye diameter is correlated with mean body depth
```{r, message=FALSE, warning=FALSE }
#now fit model
OUmodelo2<-gls(meanEye~meanBodyDepth, data=inTreeData, correlation=ou)
summary(OUmodelo2)
AIC(modelo2,OUmodelo2)
```

#### pgls to see if residual eye size is correlated with residual body depth
```{r, message=FALSE, warning=FALSE }
OUmodelo3<-gls(eye.res~depth.res, data=inTreeData, correlation=ou)
summary(OUmodelo3)
AIC(modelo3,OUmodelo3)
```

#### pgls if residual eye size is correlated with depth (avg for main text, plots of mean, min, and max for supplemental materials)
```{r, message=FALSE, warning=FALSE }
#avg depth
ou2<-corPagel(1, avgDepth.tree, form=~avgDepth.tax, fixed = FALSE)
OUmodelo4<-gls(eye.res~average.depth, data=avgDepth, correlation=ou2)
summary(OUmodelo4)
AIC(modelo4,OUmodelo4)
```

#### pgls if residual eye size is correlated with buoyancy
```{r, message=FALSE, warning=FALSE }
#avg buoyancy - proxy for water column niche use
ou3<-corPagel(1, MeanBuoy.tree, form=~MeanBuoy.tax, fixed = FALSE)
OUmodelo5<-gls(eye.res~avg.buoy, data=MeanBuoy, correlation=ou3)
summary(OUmodelo5)
AIC(modelo5,OUmodelo5)

```

### L1OU Analysis
```{r }
# l1OU Banners residual eye size, pBIC
AdjTree_resEye<-adjust_data(eye.tree, resEye)
lasso1.B<-estimate_shift_configuration(AdjTree_resEye$tree, AdjTree_resEye$Y, criterion="pBIC", lars.alg = "lasso")
lasso1.B$nShifts
#there are no shifts, otherwise the code below would be used
#lasso2.B<-estimate_convergent_regimes(lasso1.B, criterion="pBIC")
#lasso2.B$nShifts
#plot(lasso2.B)

# l1OU Banners depth, pBIC
AdjTree_depth<-adjust_data(depth.tree, as.data.frame(depth))
lasso1.depth<-estimate_shift_configuration(AdjTree_depth$tree, AdjTree_depth$Y, criterion="pBIC", lars.alg = "lasso")
lasso1.depth$nShifts
lasso2.depth<-estimate_convergent_regimes(lasso1.depth, criterion="pBIC")
lasso2.depth$nShifts
plot(lasso2.depth)

``` 

### motmot Analysis
```{r }
MMM_resEye<-matrix(resEye, dimnames=list(names(resEye)))
timeSlice.5.ml <- transformPhylo.ML(y=MMM_resEye, phy=eye.tree, model="timeSlice", 
splitTime=c(5))
outputSummary <- plot(timeSlice.5.ml, cutoff=0.001, cex=0.5, 
colour.ramp=c("blue", "red"))
```
### fitContinuous Analysis
```{r }
fitBM<-fitContinuous(eye.tree,MMM_resEye)
fitOU<-fitContinuous(eye.tree,MMM_resEye,model="OU")
fitEB<-fitContinuous(eye.tree,MMM_resEye,model="EB")
fitdelta<-fitContinuous(eye.tree,MMM_resEye,model="delta", bounds = list(delta=c( exp(-500),  5)))
fitNoise<-fitContinuous(eye.tree,MMM_resEye,model="white")

aic.vals<-setNames(c(fitBM$opt$aicc,fitOU$opt$aicc,fitEB$opt$aicc,fitdelta$opt$aicc,fitNoise$opt$aicc),
    c("BM","OU","EB", "delta", "white noise"))
aic.vals
aic.w(aic.vals)

```

